Dependent Types for Safe and Secure Web Programming

### 2013-11-11 paper/main.tex

\usepackage{amsmath}
%\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage[numbers]{natbib}
\usetikzlibrary{arrows,automata}
%\WarningsOff

\begin{document}
\fvset{fontsize=\small}
\newcommand{\idris}{\textsc{Idris}}
\newcommand{\idata}{\textsf{iData}}
\newcommand{\itasks}{\textsf{iTasks}}
\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{2013} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

@@

\title{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
%\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}

\authorinfo{Simon Fowler \and Edwin Brady}
           {School of Computer Science \\ University of St Andrews \\ St Andrews \\ Scotland}
           {Email: \{sf37, ecb10\}@st-andrews.ac.uk}

@@

\begin{abstract}
Dependently-typed languages allow for much expressive types to be used within
programs in turn allowing easier reasoning about the operation of such
programs. Stronger specifications do however bring with them the disadvantage
that it becomes increasingly difficult to write programs that are accepted by
the type checker and additional proofs may have to be specified by a user.

Embedded domain-specific languages (EDSLs) address this problem by introducing
a layer of abstraction over more specific underlying types, allowing
domain-specific code to be written in high-level languages which use dependent
types to enforce certain invariants without additional proof obligations.

In this paper, we apply this technique to web programming, and introduce an
EDSL to facilitate the creation and handling of web forms which retain type
information, reducing the scope for programmer error and attacks such as SQL
injection. We also show how EDSLs may be constructed to enforce resource usage
protocols associated with common web operations such as CGI, database access
and session handling.

@@

\section{Introduction}

Web applications, whilst ubiquitous, are also prone to incorrect construction
and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
scripting \cite{owasp:xss}. Security breaches using such exploits are
far-reaching, and high profile cases involve large corporations such as Sony,
who suffered a well-publicised and extremely costly SQL Injection breach in
2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
\cite{imperva:yahoo}.

Many web applications are written in dynamically-checked scripting languages
such as PHP, Ruby or Python, which facilitate rapid development
\cite{w3techs:webpls}. A significant drawback, however, is that such languages
do not provide  the same static guarantees about runtime behaviour afforded by
programs with more expressive, static type systems, instead relying on
extensive unit testing to ensure correctness and security.

%%------- EXAMPLE
Let us consider a simple database access routine, written in
PHP, where we wish to obtain the name and address of every employee working in
a given department, \texttt{\$dept}. We firstly construct an object
representing a database connection, where the arguments are the database host,
user, password and name respectively:

%
\begin{Verbatim}
$conn = new mysqli("localhost", "username", 
                    "password", "db");
\end{Verbatim}
%
We then check to see if the connection was successful, and exit
if not.  This check is optional, so
it would be possible to omit it: this would, however, cause
problems with later steps.

%
\begin{Verbatim}
  if (mysqli_connect_errno()) {
      exit();
  }
\end{Verbatim}
%
We then create a prepared statement detailing our query, and bind the
\texttt{`dept'} value:

\begin{Verbatim}
  $stmt = $conn->prepare("SELECT `name`, `address` 
     FROM `staff` WHERE `dept` = ?);
  $stmt->bind_param('s', $dept);
\end{Verbatim}
%
After the parameters have been bound, we execute the statement, assign
variables into which results will be stored, and fetch each row in turn. At
this point, a developer may forget to execute the statement before beginning to
fetch results or may have forgotten to bind the parameters to the statement:

%
\begin{Verbatim}
  $stmt->execute();
  $stmt->bind_result($name, $address);
  while ($stmt->fetch()) {
    printf("Name: %s, Age: %s", $name, $age);
  }
\end{Verbatim}
%
Finally, once the statement and connection are no longer needed, they should be
closed in order to discard the associated resources:

%
\begin{Verbatim}
  $stmt->close();
  $conn->close();
\end{Verbatim}
%
Even in this small example, there exists a specific resource usage protocol.
Firstly, a connection to the database must be opened. The object-oriented style
used in the example encapsulates this to an extent, as the object must be
created in order for operations to be performed, however it is less obvious in
a procedural version of the code. Secondly, a prepared statement is created,
using the raw SQL and placeholders to which variables are later bound. The
statement is then executed, and each row is retrieved from the database.
Finally, the resources are freed.

Problems may arise if the protocol is not followed correctly.
A developer may, for example, accidentally close a statement whilst still
retrieving rows, which would cause a runtime error. Similarly, a programmer may
omit closing the statement or connection, which can lead to
problems such as resource leaks in longer-running server applications.
However, in conventional programming languages, there is no way to check
automatically that a protocol is followed.

In contrast, the use of \textit{dependent types} makes it possible
to specify a program's behaviour precisely, and to check that a 
specification is followed.
%
The difficulty is 
that automatic verification by a compiler can be difficult or
often impossible, requiring additional proofs to be given by the programmer.

This complexity can be ameliorated through the use of \textit{embedded
domain-specific languages} (EDSLs), which aim to abstract away the
complexity of the underlying type system. EDSLs allow domain experts to
focus on a problem domain, rather than the specific details of their
solution.

\idris{} \cite{brady2011idris} is a language with full dependent types, and
extensive support for EDSLs through overloading and syntax macros. Through the
use of \idris{}, and a framework for describing resource protocols using
\emph{algebraic effects}~\cite{brady:effects}, we
present a dependently-typed web framework, which allows the construction of
programs with additional guarantees about correctness and security, whilst
minimising the increase in development complexity.

\subsection{Contributions}
The primary contribution of this paper is the application of 
dependent types may to provide strong static guarantees
about the correctness and security of web applications, whilst minimising
additional development complexity. In particular, we present:

\begin{itemize}
\item A form-handling mechanism, which preserves type information and
manages user input, therefore
increasing applications' resilience to attacks such as SQL injection.

\item Representations of CGI, Databases and sessions as
\textit{algebraic effects}, allowing programs to be accepted only when they
follow clearly defined resource usage protocols.

\item A practical demonstration of the framework, a message board application.

\end{itemize}

We structure the remainder of this paper as follows. We provide a brief overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain how this may be used to ensure adherence to resource usage protocols for CGI, SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for type-safe form handling in Section ~\ref{form}; and discuss the larger example of a message board system making use of these components in Section ~\ref{messageboard}.

The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/SimonJF/IdrisWeb}.
% =================================================
% =================================================

\section{An overview of the \texttt{Effects} framework}
\label{effects}
\texttt{Effects} is an EDSL which handles side-effects such as database access
as \textit{algebraic effects}, allowing greater code modularity. This section
is intended to serve as a brief introduction to the framework; for more
information, see \cite{brady:effects}.

An effect is represented as an instance of the \texttt{EFFECT} type. Such
instances may be associated with a \textit{resource}, Describing a program in
the \texttt{Effects} library is achieved using the following data type:

\begin{Verbatim}
EffM : (m : Type -> Type) ->
       (es : List EFFECT) ->
       (es’ : List EFFECT) ->
       (a : Type) -> Type
\end{Verbatim}
\texttt{EffM} is a type that is parameterised over an execution context
(\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list
of effects at the end of the program's execution (\texttt{es'}), and a return
type (\texttt{a}). The list of input types can be used to ensure that an
operation may only be called when the program is in a particular state, and
since the list of input effects may differ from the list of output effects, we
may modify the resource associated with the state after each operation. This
allows us to simply enforce a resource usage protocol, and deviation from this
protocol will result in a compile-time error.

Implementing a new effect involves creating a new algebraic data type, of kind
\texttt{Effect}, which describes the available operations and how they modify
the resource associated with the effect. Each operation consists of an initial
resource \texttt{res}, the resource at the end of the operation \texttt{res'},
and the output type of the operation \texttt{t}.

%TODO: Execution
\begin{Verbatim}
Effect : Type
Effect = (res : Type) -> (res’ : Type) ->
         (t : Type) -> Type
\end{Verbatim}

Operations are made available to users by writing functions with the
corresponding \texttt{EffM} type, which may then be used to invoke the
operation from within an effectful program.

\begin{Verbatim}
queryPostVar : 
 String -> 
 Eff m [CGI (InitialisedCGI TaskRunning)] 
   (Maybe String)
queryPostVar query = (QueryPostVar query)
\end{Verbatim}

Additionally, we define \textit{handlers}, which describe how each operation is
interpreted in a given execution context. Finally, \texttt{Effect} instances
are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff}
constructor, which is provided with the initial resource value.

% =================================================
% =================================================

\section{Modelling resource usage protocols}
\label{rup}
In this section, we show how three such effects; CGI, database access and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.

% =================================================

\subsection{CGI}
%TODO: Fix up the labels
\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]

  \tikzstyle{stateEdgePortion} = [black,thick];
  \tikzstyle{stateEdge} = [stateEdgePortion,->];
  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];

  \node[initial,state] (A)              {Uninitialised};
  \node[state]         (B) [below of=A] {Initialised};
  \node[state]         (C) [below of=B] {TaskRunning};
  \node[state]         (D) [below of=C] {TaskCompleted};
  \node[state]         (E) [below of=D] {HeadersWritten};
  \node[state]         (F) [below of=E] {ContentWritten};

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{initialise}} (B)
        (B) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startTask}} (C)
        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{finishTask}} (D)
        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeHeaders}} (E)
        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeContent}} (F);
\end{tikzpicture}
}
\caption{CGI States}
\label{fig:cgistates}
\end{figure}

CGI is used to invoke an application on a web server, making use of environment variables to convey information gained from an HTTP request and using standard output to communicate with the remote client. Importantly, HTTP headers must be correctly written to the browser prior to any other output; failure to do so will result in an internal server error.

A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell. Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program. If this were to happen, then the program would fail due to the fact that headers had not been written to the client.

By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly. In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}. This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request. We represent an uninitialised CGI process as the unit type, ().
\begin{Verbatim}[samepage]
data CGIStep = Initialised 
             | TaskRunning 
             | TaskCompleted 
             | HeadersWritten 
             | ContentWritten

data InitialisedCGI : CGIStep -> Type where
  ICgi : CGIInfo -> InitialisedCGI s
\end{Verbatim}
Figure ~\ref{fig:cgistates} shows the states through which the CGI program progresses. Each operation performed in an effectful program requires the resource to be of a certain type, and the completion of the operation may alter the type or value of the resource.

Upon creation, the CGI application is uninitialised, meaning that the environment variables and request have not been retrieved. The only operation that can be performed in this state is initialisation: by calling \texttt{initialise}, a CGIInfo record is populated, and the state transitions to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the \texttt{Cgi} effect, and involves transitioning from the uninitialised state to the initialised state.

\begin{Verbatim}
data Cgi : Effect where
    Init : Cgi () (InitialisedCGI Initialised) ()
    StartRun : Cgi (InitialisedCGI Initialised) 
                   (InitialisedCGI TaskRunning) ()
    FinishRun : Cgi (InitialisedCGI TaskRunning) 
                    (InitialisedCGI TaskCompleted) ()
    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
                       (InitialisedCGI HeadersWritten) ()
    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
                       (InitialisedCGI ContentWritten) ()
    OutputData : String -> 
                 Cgi (InitialisedCGI TaskRunning) 
                     (InitialisedCGI TaskRunning) ()
    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
                               :: effs) -> 
                       CGIProg effs a -> 
                       Cgi (InitialisedCGI TaskRunning) 
                           (InitialisedCGI TaskRunning) a
\end{Verbatim}
Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.

User code executes in the \texttt{TaskRunning} state. Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers. It is important to note, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record. This data may then be printed at the end of the program's execution, in accordance with the resource usage protocol.

After the user code has finished execution, control returns to the library code. At this point, the state transitions to TaskCompleted, the headers are written, and the headers are output to the request. Finally, the headers and content are written which completes the process. Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.

In \idris{}, types are first-class, meaning that they may be treated like other terms in computations. We may therefore define the following type synonym, which we may use in order to make use of the CGI framework: 
\begin{Verbatim}
CGIProg : List EFFECT -> Type -> Type
CGIProg effs a = 
 Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a
\end{Verbatim}
This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.

A simple ``Hello, world!'' program would be defined as follows:

\begin{Verbatim}
module Main
import Cgi

sayHello : CGIProg [] ()
sayHello = output "Hello, world!"

main : IO ()
main = runCGI [initCGIState] sayHello
\end{Verbatim}
% =================================================

\subsection{Database access with SQLite}

%TODO: Fix up the labels
\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzset{every loop/.style={min distance=10mm,looseness=5}}
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]

  \tikzstyle{stateEdgePortion} = [black,thick];
  \tikzstyle{stateEdge} = [stateEdgePortion,->];
  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];

  \node[initial,state] (A)              {Uninitialised};
  \node[state]         (B) [below of=A] {ConnectionOpened};
  \node[state]         (C) [below of=B] {PreparedStatementOpen};
  \node[state]         (D) [below of=C] {PreparedStatementBinding};
  \node[state]         (E) [below of=D] {PreparedStatementBound};
  \node[state]         (F) [below of=E] {PreparedStatementExecuting};

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{openDB}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, text width=2cm, xshift=3em] {\texttt{closeDB, connFail}} (A)
        (B) edge[stateEdge]   node[edgeLabel, xshift=4em] {\texttt{prepareStatement}} (C)
        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=3em, yshift=-4em] {\texttt{stmtFail}} (A)
        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startBind}} (D)
        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{finishBind}} (E)
        (E) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{bindFail}} (A)
        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{beginExecution}} (F)
        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=-3.5em] {\texttt{finalise}} (B)
        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{executeFail}} (A)
            edge[stateEdge, loop left]   node[edgeLabel, text width=2cm, xshift=2em] {\texttt{nextRow, reset}} (B);
\end{tikzpicture}
}
\caption{Database Resource Usage Protocol}
\label{fig:sqlitestates}
\end{figure}

SQLite\footnote{\texttt{http://www.sqlite.org}} is a lightweight SQL database engine often used as simple, structured storage for larger applications. We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems.

The creation, preparation and execution of SQL queries has a specific usage protocol, with several possible points of failure. Failure is handled in traditional web applications by the generation of exceptions, which may be handled in the program.
Handling such exceptions is often optional, however, and in some cases unhandled errors may cause a deployed web application to display an error to the user. Such errors can be used to determine the structure of an insecure SQL query, and are often used by attackers to determine attack vectors for SQL injection attacks.

Figure ~\ref{fig:sqlitestates} shows a resource usage protocol for database access, which we have implemented for the SQLite library. This is encapsulated by the \texttt{Sqlite} effect; we again omit some operations, such as those to bind and retrieve data types other than \texttt{String}, in the interest of brevity.

\begin{Verbatim}[samepage]
data Sqlite : Effect where
  OpenDB : 
    String -> 
    Sqlite () 
           (SQLiteRes ConnectionOpened) Bool
  CloseDB : 
    Sqlite (SQLiteRes ConnectionOpened) () Bool
  PrepareStatement : 
    String -> 
    Sqlite (SQLiteRes ConnectionOpened) 
           (SQLiteRes PreparedStatementOpen) Bool
  StartBind : 
    Sqlite (SQLiteRes PreparedStatementOpen) 
           (SQLiteRes PreparedStatementBinding) ()
  BindText : 
    ArgPos -> String -> Int -> 
    Sqlite (SQLiteRes PreparedStatementBinding) 
           (SQLiteRes PreparedStatementBinding) 
           Bool
  FinishBind : 
    Sqlite (SQLiteRes PreparedStatementBinding) 
           (SQLiteRes PreparedStatementBound) 
           Bool
  ExecuteStatement : 
    Sqlite (SQLiteRes PreparedStatementBound) 
           (SQLiteRes PreparedStatementExecuting) ()
  Finalise : 
    Sqlite (SQLiteRes PreparedStatementExecuting) 
           (SQLiteRes ConnectionOpened) 
           Bool
  GetColumnText : 
    Int -> 
    Sqlite (SQLiteRes PreparedStatementExecuting)
           (SQLiteRes PreparedStatementExecuting)
           String
  RowStep : 
    Sqlite (SQLiteRes PreparedStatementExecuting)
           (SQLiteRes PreparedStatementExecuting)
           StepResult
\end{Verbatim}
There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query. This is reflected in the associated resource \texttt{SQLiteRes}, which again is parameterised by the current protocol state.
\begin{Verbatim}
data SQLiteRes : Step -> Type where
  OpenConn : DBPointer -> SQLiteRes s
  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
  ExecutingStmt : DBPointer -> 
                  StmtPointer -> 
                  StepResult -> 
                  SQLiteRes s

data DBPointer = ValidConn Ptr
               | InvalidConn

data StmtPointer = ValidStmt Ptr
                 | InvalidStmt 
\end{Verbatim}
If a failure happens at any point during the computation, the resource is updated to reflect the failure: if, for example, the library failed to create a connection to the database, the resource value would be updated to \texttt{OpenConn InvalidConn}. At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety. The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial protocol state.

%Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
SQL queries are evaluated in SQLite upon a call to the C library function \texttt{sqlite3\_step()}. In the case that a statement returns a result set, each subsequent call retrieves another row for processing using a column access function. Once all rows have been retrieved, the library returns \texttt{SQLITE\_DONE}, meaning that no further calls should be made without resetting the function. We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor. 
\begin{Verbatim}
data StepResult = Unstarted
                | StepFail
                | StepComplete
                | NoMoreRows
\end{Verbatim}
Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} C library function, returns a result of type \texttt{StepResult}, which is then reflected in the resource. Calls to \texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult} is either \texttt{Unstarted} or \texttt{StepComplete}. We may therefore statically guarantee that only calls that will return a valid result are executed.

By incorporating pointers to open connections and prepared statements into the resource associated with the effect, we introduce a further layer of abstraction, in order to hide implementation details from the developer and encourage cleaner, less error-prone code.

\subsubsection{Example Code}
Programs making use of the DSL should look familiar to developers even without a background in functional programming. To demonstrate the functionality of the library, we return to the previous example of selecting the names and addresses of all staff working in a given department. Due to the fact that the \texttt{Effects} library overloads the bind operator, we may make use of \texttt{do}-notation, facilitating the usage of an imperative style.

We define a function of type:
\begin{Verbatim}
String ->
Eff IO [SQLITE ()] (Either String (List (String, String)))
\end{Verbatim}
This means that the program will be run in the IO execution context, and must start and end with no active resources. The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned. 
\begin{Verbatim}
testSelect : String -> Eff IO [SQLITE ()] 
             (Either String (List (String, Int)))
testSelect dept = do
  open_db <- openDB "people.db"
  if open_db then do
    let sql = "SELECT name, address FROM `staff` 
                    WHERE dept = ?;"
    sql_prep_res <- prepareStatement sql
    if sql_prep_res then do 
      startBind
      bindText 1 dept
      finishBind
      beginExecution
      results <- collectResults
      finaliseStatement
      closeDB
      return $ Right results
    else do err <- stmtFail
            return $ Left err
  else do err <- connFail
          return $ Left err 
\end{Verbatim}
The program initially attempts to open a connection to the \texttt{people.db} database. At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state. The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful. If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.

A call to \texttt{prepareStatement} attempts to create a prepared statement, and a subsequent call to \texttt{beginExecution} allows data to be retrieved from the database.

\texttt{CollectResults} is a simple function which makes a call to \texttt{nextRow} in order to make the next row of the result set available for processing, and uses the \texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data from the database. This function is then recursively called until there are no more rows to process.
\begin{Verbatim}
collectResults : 
  Eff IO 
  [SQLITE (SQLiteRes PreparedStatementExecuting)] 
  (List (String, Int))
collectResults = do
  step_result <- nextRow
  case step_result of
      StepComplete => do name <- getColumnText 1
                         age <- getColumnInt 2
                         xs <- collectResults
                         return $ (name, age) :: xs
      NoMoreRows => return []
      StepFail => return [] 
\end{Verbatim}
In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process. In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
\begin{Verbatim}
data DBVal = DBInt Int
           | DBText String
           | DBFloat Float
           | DBNull
\end{Verbatim}
We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
\begin{Verbatim}
  ResultSet : Type
  ResultSet = List (List DBVal)
\end{Verbatim}
Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
\begin{Verbatim}
executeSelect : 
  String -> String -> List (Int, DBVal) -> 
  (Eff m [SQLITE (SQLiteRes PreparedStatementExecuting)] 
                 (List DBVal)) -> 
  Eff m [SQLITE ()] (Either String ResultSet)
\end{Verbatim}
% =================================================
\subsection{A Simple Session Handler}
More complex web applications require some persistent state across separate requests. This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data. In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.

A major strength of the \texttt{Effects} library is that it allows for simple composition of individual, fine-grained effects. By combining the individual CGI and SQLite components, we may construct a simple session handler to provide a notion of state across separate web requests.

We implement this by having a SQLite database, containing two tables: \texttt{session}, which stores session keys and their associated expiry dates, and \texttt{sessiondata}, which contains the data associated with each session. A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.
%TODO: Fix up the labels
\begin{Verbatim}[samepage]
data Session : Effect where
  LoadSession : 
  SessionID -> 
  Session (SessionRes SessionUninitialised) 
          (SessionRes SessionInitialised) 
          (Maybe SessionData)
  UpdateSession : 
    SessionData -> 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) ()
  CreateSession : 
    SessionData -> 
    Session (SessionRes SessionUninitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionID)
  DeleteSession : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) Bool 
  WriteToDB : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) Bool
  DiscardSessionChanges : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) ()
  GetSessionID : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionID)
  GetSessionData : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionData)
\end{Verbatim}
\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]

  \tikzstyle{stateEdgePortion} = [black,thick];
  \tikzstyle{stateEdge} = [stateEdgePortion,->];
  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];

  \node[initial,state] (A)              {SessionUninitialised};
  \node[state]         (B) [below of=A] {SessionInitialised};

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
\end{tikzpicture}
}
\caption{Session Handler Resource Usage Protocol}
\label{fig:sessionstates}
\end{figure}
Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated with the session handler. In this application, there exist two states: \texttt{SessionUninitialised}, in which the user may load or create a new session, and \texttt{SessionInitialised}, in which the user may update the representation of the session in memory, serialise the session and write it to the database, or delete the session and invalidate the user's session key. The introduction of these two states ensures that changes are explicitly either written or discarded, eliminating the possibility of a developer updating the session but neglecting to commit it to persistent storage. This, of course, is under the assumption that the process exits cleanly: we attempt to facilitate this by writing total functions where possible.

Much like the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
\begin{Verbatim}
data SessionStep = SessionUninitialised
                 | SessionInitialised

data SessionRes : SessionStep -> Type where
  InvalidSession : SessionRes s  
  ValidSession   : SessionID -> 
                   SessionData -> 
                   SessionRes s
\end{Verbatim}
The \texttt{SessionRes} data type is parameterised over the current state, which determines which operations may be performed, and has two constructors: \texttt{InvalidSession} and \texttt{ValidSession}. If an operation such as creating a new session fails, no further side-effecting calls will be made, in turn preserving integrity.

% =================================================

\section{Type-aware form handling}
\label{form}
Programming web applications often involves processing user data, which may then be used in further effectful computations. Data submitted using a form is transmitted over the internet as a string as part of an HTTP request, which traditionally involves losing associated type information.

This can in turn increase lead to risks; developers may assume that a data is of a certain type, and therefore discount the possibility that it may have been modified by an attacker. One example would be the traversal of paginated data, in which a form is used to make a request to retrieve the next page of data. This may involve sending an integer detailing the current page, which could be used in a query such as:

\begin{Verbatim}
'SELECT `name`, `address` FROM `staff` LIMIT ' + 
       page + ', 5';
\end{Verbatim}
The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT clauses here

In this section, we present a mechanism by which we introduce a web form DSL for the creation of forms which preserve type information. Once the form has been submitted, retrieved information is passed directly to a developer-specified function for handling, without the need to manually check and deserialise data.

We begin with a simple example of a form which requests a user's name, and echoes it back. Firstly, we define a form handler which echoes back a string provided by the form handler. It has one argument of type \texttt{Maybe String}, which accounts for the possibility that the user may have specified incorrect data within the form.

\begin{Verbatim}
sayHello : Maybe String -> 
           FormHandler [CGI (InitialisedCGI TaskRunning)]
sayHello (Just name) = output ("Hello, " ++ name ++ "!")
sayHello _ = output "Error!"
\end{Verbatim}

We then specify this in a list of handlers, detailing the arguments, available effects, handler function and unique identifier:

\begin{Verbatim}
handlers : HandlerList
handlers = [(handler args=[FormString], 
                     effects=[CgiEffect], 
                     fn=sayHello, 
                     name="sayHello")
           ]
\end{Verbatim}

We also define a form to take in a name from the user, and specify that it should use the \texttt{sayHello} handler.

\begin{Verbatim}
showHelloForm : UserForm
showHelloForm = do
  addTextBox "Name" FormString Nothing
  useEffects [CgiEffect]
  addSubmit sayHello handlers
\end{Verbatim}

Finally, we specify that if data has been submitted for processing, then it should be passed to the form handler. If not, then the form should be shown.

\begin{Verbatim}
cgiHello : CGIProg [] ()
cgiHello = do
  handler_set <- isHandlerSet
  if handler_set then do
    handleForm handlers
    return ()
  else do
    addForm "nameform" "helloform" showHelloForm
    return ()

main : IO ()
main = runCGI [initCGIState] cgiHello
\end{Verbatim}
When this CGI application is invoked, it will begin by outputting a form to the page, requesting a name from the user. Upon submission of the form, the form handler will be invoked, and the name will be used in the output.

>>>>

\usepackage{amsmath}
%\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage[numbers]{natbib}
%\WarningsOff

\newcommand{\useverbtb}[1]{\begin{tiny}\BUseVerbatim{#1}\end{tiny}}
\newcommand{\useverb}[1]{\begin{small}\BUseVerbatim{#1}\end{small}}
\newcommand{\useverbb}[1]{\begin{small}\BUseVerbatim{#1}\end{small}}

\begin{document}
\fvset{fontsize=\small}
\newcommand{\idris}{\textsc{Idris}}
\newcommand{\idata}{\textsf{iData}}
\newcommand{\itasks}{\textsf{iTasks}}
\newcommand{\Idris}{\textsc{Idris}}
\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{IFL '13}{August 28--31, 2013, Nijmegen, The Netherlands} 
\copyrightyear{2013} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

@@

\title{Dependent Types for Safe and Secure Web Programming}
\authorinfo{Simon Fowler \and Edwin Brady}
           {School of Computer Science, University of St Andrews, St Andrews, Scotland}
           {Email: \{sf37, ecb10\}@st-andrews.ac.uk}

@@

\begin{abstract}
Dependently-typed languages allow precise types to be used during development,
facilitating easier reasoning about the operation of programs. However,
stronger types bring a disadvantage that it becomes increasingly difficult to
write programs that are accepted by a type checker and additional proofs may
have to be specified by a programmer.

Embedded domain-specific languages (EDSLs) address this problem by introducing
a layer of abstraction over more specific underlying types, allowing
domain-specific code to be written in a high-level languages which uses
dependent types to enforce certain invariants without additional proof
obligations.

In this paper, we apply this technique to web programming.  Using the
dependently typed programming language \Idris{}, we introduce an EDSL to
facilitate the creation and handling of \emph{statically checked} web forms,
reducing the scope for programmer error and attacks such as SQL injection. We
also show how to enforce resource usage protocols associated with common web
operations such as CGI, database access and session handling.

@@

\input{introduction}
\input{effects}
\input{protocols}
\input{forms}
\input{messageboard}
\input{conclusion}
